Extended Data

### 2021-04-28 doc/manuscript/Dow_manuscript_Nature.Rmd
authors/ affiliations, formatting cleanup

---
title: ''
output:
  word_document
  #bookdown::word_document2: 
    #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #fig_caption: yes
bibliography: references.bib
csl: nature.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

*Grant Harley*

@@

Ian McGregor^1,5^

@@

Sean McMahon^6,7^

@@

Kristina J. Anderson-Teixeira^1,7^*

@@

5. NCSU

6. SERC

7. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests (Fig. `r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America.** 
First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[*D'Orangeville*], and set in the context of leaf phenology at these sites.

```{r schematic_summary, out.width="100%", echo=FALSE, fig.align="center", fig.cap= schematic_summary_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/schematic_summary.png")
```

@@

```{r pheno_Tsensitivity, out.width="80%", echo=FALSE, fig.align="center", fig.cap=DOY_Tsensitivity_caption}
# Created in RScripts/bert/wood_phenology_analysis_bert.R
knitr::include_graphics("tables_figures/DOY_Tsensitivity.png")
```

```{r DOYtiming, out.width="100%", echo=FALSE, fig.align="center", fig.cap= DOYtiming_caption}
# Created in RScripts/additional_figures.R
knitr::include_graphics("tables_figures/DOYtiming_allyears.png")
```

@@

```{r tree_rings, out.width="100%", echo=FALSE, fig.align="center", fig.cap= tree_rings_caption}
knitr::include_graphics("tables_figures/tree_rings_tmax.png")
```

@@

**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2`.** 
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 
*(describe results across all sites)*
(Fig. `r tree_rings`)

@@

# Methods

## Study sites

@@

## Data collection and preparation
### Dendrometer bands

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).

@@

### Climate
For the dendrometer band analysis, we used climate data from local weather stations. 
Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020]. 
Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
The R package climpact () was used to plot temperatures for visual inspection. 
Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
(*how were gaps filled for Harvard? Or was the record good?*)

@@

## Determining critical temperature windows

@@

## Hypothesis testing

@@

# Acknowledgements
people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
Funding: ForestGEO

# Author Contributions

# Additional Information

@@

# References

>>>>

---
title: ''
output:
  bookdown::word_document2: 
    reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    fig_caption: yes
bibliography: references.bib
csl: nature.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Grant Harley

@@

Ian McGregor^1,6^

@@

Sean McMahon^7,9^

@@

Alan Tepley^1,8^

Kristina J. Anderson-Teixeira^1,9^*

@@

5. Department of Geography & Geological Sciences, University of Idaho, ID, USA

6. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

7. Smithsonian Environmental Research Center, Edgewater, MD, USA

8. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

9. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
However, less is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood that has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

**Here, we characterize how early spring temperatures affect stem growth phenology, growth rates, and annual growth of temperate deciduous trees in two forests (Fig. `r schematic_summary`) and annual growth across these and an additional `r n_cores_sites-2` forests spanning eastern North America (`r EDF_map`).** 
First, using dendrometer band measurements taken throughout the growing season at the Smithsonian Conservation Biology Institute (SCBI; Virginia; n = **146** trees over 10 years) and Harvard Forest (Massachusetts; n = **755** trees over six years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- and diffuse-porous species, which differ in growth phenology[*D'Orangeville*], and set in the context of leaf phenology at these sites.

![`r schematic_summary_caption`](tables_figures/schematic_summary.png){ width=100% }

@@

![`r DOY_Tsensitivity_caption`](tables_figures/DOY_Tsensitivity.png){ width=50% }

![`r DOYtiming_caption`](tables_figures/DOYtiming_allyears.png){ width=50% }

@@

![`r tree_rings_caption`](tables_figures/tree_rings_tmax.png){ width=100% }

@@

**To set these findings in a broader scale context, we analyzed an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` (`r EDF_map`).** 
In total, our analysis included # species at sites distributed from XXX to XXX and spanning a ## degree range in April T.** 
*(describe results across all sites)*
(Fig. `r tree_rings`)

@@

# Methods {-#Methods}

## Study sites {-#Study}

@@

## Data collection and preparation {-#Data} 
### Dendrometer bands {-#Dendrometer}

@@

### Tree rings {-#Tree} 
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).

@@

### Climate {-#Climate} 
For the dendrometer band analysis, we used climate data from local weather stations. 
Daily maximum and minimum temperatures ($T_{max}$ and $T_{min}$, respectively) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot, via the ForestGEO Climate Data Portal v1.0 [https://forestgeo.github.io/Climate/; @anderson-teixeira_forestgeo_2020]. 
Daily temperature records for Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).
The R package climpact () was used to plot temperatures for visual inspection. 
Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. 
Gaps in the SCBI meteorological tower data were subsequently filled using temperature readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). 
(*how were gaps filled for Harvard? Or was the record good?*)

@@

## Determining critical temperature windows {-#Determining}

@@

## Hypothesis testing {-#Hypothesis}

@@

# Acknowledgements {-#Acknowledgements} 
people: (everyone who helped collect data-- see list on GitHub), Jess Shue (data)
Funding: ForestGEO

# Author Contributions {-#Author}

# Additional Information {-#Additional}

@@

# References {-#References}
